A note on observation processes in epidemic models

### 2019-08-15 ms/ms.tex
tweaks

Epidemic models provide valuable tools for understanding how a disease
spreads in a population. Many epidemic models focus on characterizing
the underlying transmission mechanism but make naive assumptions about
how infections are reported. In this note, we use a simple deterministic
Susceptible-Infected-Removed (SIR) model to compare two common assumptions 
about disease incidence reports: individuals can report their infection as soon as
they become infected or as soon as they recover. We show that making
wrong assumptions about the underlying observation processes can result
in biased estimates of the basic reproduction number and narrow confidence
intervals.

@@

Despite the amount of effort put into developing \emph{better} process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Many disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or develops symptom (e.g., 
\cite{bhadra2011malaria, king2015avoidable}); 
some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious).

@@

First, we compare the deterministic dynamics of incidence curves,
$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
the observed epidemic peak timing and reduces the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
might not expect to be able to distinguish between the two reporting processes
based on the time series alone.

Note that incidence is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the observed dynamics of incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While they are relatively
uncommon, some models do not make a clear distinction between a discrete-time observed 
variable and a continuous-time latent variable (Table 1).

\begin{table}
\centering
\begin{tabular}{l | c | c}
\hline
\textbf{Study} & \textbf{Disease} &  \textbf{Observed variables}\\
\hline
\cite{capistran2009parameter} & Respiratory syncytial virus & Number of hospital cases\\
\cite{hooker2010parameterizing} & Measles & Number of reported cases\\
\cite{eisenberg2013identifiability} & Cholera & Number of reported cases\\
\cite{yang2013stability} & Hand-foot-mouth disease & Number of reported cases\\
\cite{gonzalez2014fractional} & Influenza & Number of confirmed cases\\
\hline
\end{tabular}
\caption{
\textbf{List of studies that fit prevalence curves to incidence data.}
Incidence, defined as the number of new cases between two reporting periods, depends
on the reporting time step.
Prevalence, defined as the number of currently infected individuals, does not depend
on the reporting time step.
Some models do not make a clear distinction between these two quantities.
}
\end{table}

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
bias changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

@@

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population; however, many epidemic models make 
naive assumptions about how cases are reported. 
Here, we used a deterministic SIR model to show that 
making wrong assumptions about observational processes in epidemic models
can give biased estimates of the basic reproduction number and narrow
confidence intervals.

@@

We considered observation processes associated with incidence 
reporting; however, we expect observation processes to be just as important
in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
These frameworks do not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: delays in case reports can decrease
the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1).

@@

Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of infectious disease transmission.
We caution disease modelers to be more mindful about their decisions in developing
epidemic models and the implications of their model assumptions.

>>>>

Epidemic models provide valuable tools for understanding how a disease
spreads in a population. Many disease modelers focus on characterizing
the underlying transmission mechanism but make naive assumptions about
how infections are reported. In this note, we use a simple deterministic
Susceptible-Infected-Removed (SIR) model to compare two common assumptions 
about disease incidence reports: individuals can report their infection as soon as
they become infected or as soon as they recover. We show that making
wrong assumptions about the underlying observation processes can introduce
small bias to the estimates of the basic reproduction number and give
narrow confidence intervals.

@@

Despite the amount of effort put into developing \emph{better} process models, 
disease 
modelers often neglect details of the observation process associated with disease
case reports. Many disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or when an individual becomes symptomatic
(e.g., \cite{bhadra2011malaria, king2015avoidable}); 
some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious).

@@

First, we compare the deterministic dynamics of incidence curves,
$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
the observed epidemic peak timing and reduces the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
\emph{might} not expect to be able to distinguish between the two reporting processes
based on the time series alone.

Note that incidence is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the observed dynamics of incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While they are relatively
uncommon, some models do not make a clear distinction between prevalence and 
incidence 
\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in a clearer bias (in opposite directions) 
and a much lower coverage (Figure 2B).

@@

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population; however, epidemic models often make 
simple assumptions about how cases are reported. 
Here, we used a deterministic SIR model to show that 
making wrong assumptions about observational processes in epidemic models
can give biased estimates of the basic reproduction number and narrow
confidence intervals.

@@

We considered observation processes associated with incidence 
reporting; however, we expect observation processes to be just as 
(if not more) important in analyzing mortality data. In particular,
many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
These frameworks do not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve -- delays in case reports can decrease
the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1).

@@

Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of infectious disease transmission.
While the amount of bias introduced from the misspecification of the observation
model can be fairly small (e.g., less than 5\% in our examples), any modelers
trying to make serious forecasts should be aware of potential biases.
We caution disease modelers to be more mindful about their decisions in developing
epidemic models and the implications of their model assumptions.
